Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-18 paper.tex
shortend bib, reworked some parts, intro FIRL as ac, moved fig 4

\and
%\linebreakand %\and
\IEEEauthorblockN{Mirko Viroli}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
Alma Mater Studiorum — Università di Bologna}\\
Cesena, Italy\\
mirko.viroli@unibo.it % 0000−0003−2702−5702
}
\and
\IEEEauthorblockN{Lukas Esterle}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
Aarhus University}\\
Aarhus, Denmark\\
lukas.esterle@ece.au.dk}
}
%
\maketitle              % typeset the header of the contribution
%
\input{scala.tex}
\lstset{language={scafi}}
%%% Comment command, to be removed before submission
\newcommand{\meta}[3]{\textcolor{#1}{\textbf{#2}: #3}}
\newcommand{\ga}[1]{\meta{red}{GA}{#1}}
\newcommand{\lukas}[1]{\meta{purple}{Lukas}{#1}}
\newcommand{\mv}[1]{\meta{green}{MV}{#1}}
%% ACRONYMS
\acrodef{DecPOMDP}[DecPOMDP]{decentralized partially observable Markov decision processes}
\acrodef{RL}[RL]{reinforcement learning}
\acrodef{MARL}[MARL]{multi-agent reinforcement learning}
\acrodef{MAARL}[ManyRL]{many-agent reinforcement learning}
\acrodef{MDP}[MDP]{Markov Decision Process}
\acrodef{MLP}[MLP]{multi-layer perceptron}
\acrodef{GNN}[GNN]{graph neural network}
\acrodef{CNN}[CNN]{convolutional neural network}
\acrodef{CTDE}[CTDE]{centralized training and decentralized execution}
\acrodef{DQN}[DQN]{Deep Q-Learning}
%\ga{Page limit: 10 pages (include refs)}

@@

In this context, two main approaches have been proposed for creating distributed controllers out of macro-level task/goal descriptions: 
 \emph{manual design}, in which programmers build the controllers directly, and 
 \emph{automatic design}, which involves synthesizing programs using machine learning methods.
%
In this paper, we consider a new \emph{hybrid} approach called  \emph{Field-Informed reinforcement learning}. We utilise manually designed \emph{computational fields} (globally distributed data structures) to manage the interaction between agents. Using \emph{Deep Q-learning} in combination with \emph{Graph Neural Networks} we enable the agents to learn the necessary local behaviour automatically to solve collective tasks. 
%
This allows us to create distributed controllers informed by a collective knowledge 
 that has been distilled during learning, but that uses only local information at runtime.
% 
We demonstrate the effectiveness of this new approach in simulated use cases 
 where tracking and covering tasks for swarm robotics are successfully solved. 
\end{abstract}

@@

%This leads to several questions 
%\begin{enumerate}
%	\item How to distribute sensors to ensure good attainment of information?
%	\item How to maintain knowledge about the phenomena and unlearn irrelevant information?
%	\item How to move sensors when the phenomena moves to ensure good coverage?
%\end{enumerate} \lukas{refine and rework these questions}
The coordination of a group of autonomous agents that can perceive and act in their environment is a fundamental problem in artificial intelligence. 
Such agents need to cooperate and communicate in order to achieve a common goal, 
 while dealing with the challenges of distributed and situated intelligence. 
 These challenges include the limited and local nature of the information available to each agent and the emergence of global behaviour from local interactions. 
%
One domain where these challenges are particularly evident is Cyber-Physical Swarms (CPSs, or \emph{swarm-like systems}), 
 where a large number of agents interact with each other and their environment based on spatial proximity. 
%
Examples of CPSs include swarm robotics, smart cities, sensor networks, and social systems. 
 In these systems, agents need to adapt to dynamic and uncertain situations, while ensuring the achievement of global objectives that may not be directly observable or measurable by individual agents.

A key question in designing swarm-lie systems is how to create distributed controllers for the agents that can enable them to perform complex collective tasks. 
 Two main approaches have been proposed for this problem: manual design and automatic design. 
%
In manual design, programmers build the controllers directly, 
 using domain knowledge and programming languages or frameworks that support distributed computation and communication like macro-programming approaches~\cite{DBLP:journals/corr/abs-2201-03473}. 
%
In automatic design, machine learning methods like \ac{MARL} and evolutionary algorithms are used to synthesize programs or policies for the agents from the high-level task or goal descriptions.
%
Both approaches have advantages and disadvantages. 
 Manual design allows programmers to specify desired properties for the system declaratively. However, manual design can be tedious and error-prone.
%
Automatic design can overcome these limitations by learning from data and experience. 
 However, automatic design can also suffer from several challenges, such as learning the right representation for agent communication.
%

In this paper, we propose a novel hybrid approach that combines manual design and automatic design of distributed controllers.
This follows a novel trend in which high-level declarative programming languages are mixed with machine learning techniques in order to synthesise robust collective controllers~\cite{DBLP:conf/acsos/Aguzzi21,DBLP:conf/icdcs/AguzziCV22,DBLP:conf/coordination/AguzziCV22}.
Specifically, we propose a solution called field-informed reinforcement learning (FIRL) that utilizes aggregate computing~\cite{Beal2015Computer} together with a \ac{GNN}~\cite{Zhou2020AIOpen} in combination with a reinforcement learning approach, namely \ac{DQN}~\cite{mnih2015human}. 
 First, aggregate computing is utilised to collect information from neighbours and create a local knowledge base for each agent. Second, we utilise a \ac{GNN} within the \ac{DQN}. 
 Here the \ac{GNN} provides node embeddings for each agent, serving as input for the \ac{DQN}. 
 The \ac{DQN} provides then the appropriate actions for the agent to achieve their tasks.
%We further combine this with reinforcement learning techniques to respond to changing conditions in the environment. 
%
The main contribution of FIRL is that it allows us to create distributed controllers that are informed by a collective knowledge that has been distilled during learning, but that use only local information at runtime. This way, FIRL can achieve a balance between manual design and automatic design, combining the benefits of both approaches while mitigating their drawbacks.
%% Perhaps to much detail 
FIRL can also be seen as a way of bridging the gap between symbolic and sub-symbolic AI methods, by integrating declarative programming with deep learning.

@@

\section{Background and Motivation}
\label{sec:background}
In this section, we aim to provide a concise overview of the system we intend to address, 
 starting with a high-level system model. 
%
We then proceed to discuss relevant research areas that inform our approach. 
%
Finally, we present a formalization of the problem and a discussion of the motivations behind our proposed work.
%\ga{Page budget: 1/2 page \\}
%\ga{Plan: discussion about the problem of coordination in multi-agent systems and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{Swarm systems}
This article studies intelligent \emph{collective behaviours} within the context of \emph{large-scale} distributed systems.
 Specifically, we focus on \emph{Cyber-Physical Swarms} or \emph{swarm-like systems}, 
 i.e.,  systems consisting of \emph{devices} (or nodes) equipped with \emph{sensors} and \emph{actuator} able to interact with the real world.
 Each of these nodes can interact with its direct neighbours, either based on physical (e.g. communication range) or logical distance (e.g. 1-hop neighbourhood).

%
Each node is governed by an autonomous software agent, 
 executing a \emph{local} control loop. 
 In each iteration of this loop, the agent can access the information available within its own \emph{context}. 
 This context is comprised of information acquired by the node directly through its sensors and the information received from its neighbours. 
%
At the end of each control loop iteration, 
 the node can send messages to its neighbourhood. 
 This message may contain raw sensor data or aggregated information over time or from other neighbours.
% about its current evaluation and update its internal state.
%
%This general model can be adapted to several different scenarios, such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
Examples of such large-scale systems can be a network of robots (e.g., swarm robotics), camera and IoT networks, or a network of mobile phones.
%
Our goal is to find a \emph{homogeneously} distributed controller $\pi$ (i.e., the same controller for the whole system) that, 
 starting from \emph{only} local configurations, leads 
 the system to achieve a certain \emph{collective} requirements through \emph{cooperation}, 
 such as spatial area coverage, phenomena tracking, and robot aggregation~\cite{DBLP:journals/firai/SchranzUSE20}.
%
The homogeneity of the controller is a key requirement to ensure the \emph{scalability} of the approach, 
 as it allows us to avoid the need for a centralised controller that would be a bottleneck for the system,
 and it is the typical choice in swarm-like systems~\cite{brambilla2013swarm,yang2021many,pmlr-v80-yang18d,DBLP:conf/aaai/ZhengYCZZWY18}.
%
\subsection{Field-based Coordination}
%\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
The field-based coordination approaches utilise a concept of \emph{computational fields} 
 (or simply \emph{fields}), 
 which are \emph{distributed} data structures that associate locations with values and evolve over time. 
%
\emph{Aggregate computing}~\cite{Beal2015Computer}, a modern field-coordination approach, is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 
% 
This \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm can be realised to devise collective and self-organizing behaviour through a composition of functions operating on fields. 
%
The fields map a set of individual agents to computational values, 
 allowing them to associate what they sense, process, and actuate. 
%
Fields are computed locally but subject to a global viewpoint, 
 enabling emergent collective behaviour through the interplay of the \emph{system model} (i.e., structure and dynamics of the systems)
 and the \emph{programming model} (i.e., how collective behaviours are expressed) known as field calculus~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}.
%
The system model is structurally similar to \emph{swarm-like systems}, 
 as it is composed of a set of \emph{agents} 
 that interact with each other through a \emph{network} of \emph{neighbourhood relations}.
%
The dynamics of each agent follow a local control loop called \emph{round}
 that comprises the following steps:
\begin{enumerate}
  \item \emph{Sense}: the agent acquires information from the environment and collects the messages of its neighbourhood, building a \emph{context};
  \item \emph{Compute}: the agent executes a \emph{program} that computes an \emph{export}, that is a set of values to be sent to the neighbourhood;
  \item \emph{Act}: the agent sends the export to its neighbourhood and updates its internal state acting on the environment.
\end{enumerate}
The proactive and iterative execution of the round by each agent leads to the emergence of collective behaviour specified by the program.

@@

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source node and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. 
%
In ScaFi~\cite{casadei2022scafi} (a scala implementation of aggregate computing) it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the field expansion. 
This pattern is called \emph{gradient-cast} (or \emph{G}) and is a fundamental building block for the coordination of swarm-like systems.
It is expressed as: 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
%\ga{more description here}
where \texttt{source} is a Boolean field that indicates the presence of a source node, 
 \texttt{value} is the value to be cast, and 
 \texttt{acc} is the accumulation function that is applied to the value during the cast.

\lukas{Lukas to review this section again.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $f_v$.
For the sake of simplicity, 
 we thereafter describe $G_f$ as a graph that contains the feature set $f_v$ for each node $v \in V$.
 Note that, when we refer to $G_f$ and $G_o$ we are referring to the same graph $G$ but with different node features.
 Also, to access the feature set $f_v$ of a node $v \in V$ we use the notation $f_v$ or $G_f[v]$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $f_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. 
 The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
%
In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $\mathcal{N}_G(v)$,
 and then combining it with the node's current embedding $h_v$.
%
The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. 
 With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}_G(v)\right\}\right)   
\end{equation}

@@

  \includegraphics[width=\linewidth]{imgs/architecture.pdf}

@@

  \caption{High-level description of Field-Informed reinforcement learning dynamics. 

@@

  Using this feature, an aggregate program computes spatio-temporal information that enhances the feature set of each node, depicted as colours in the middle graph. 
  Finally, utilizing the GNN, actions are computed for each node in the system to be performed against the environment, enabling advancement in the simulations according to swarMDP rules.

@@

 representing the participating nodes and their neighbourhood relations at time $t$, as input. 

@@

The policy $\pi_{gnn}$ is then evaluated for each node using $f_v$ as input, 

@@

The individual nodes only combine information from their local neighbourhood.

@@

In fact, the program $\Gamma$ is proactively executed at every node, 

@@

otherwise, they could not have a local interpretation for each node, according to our system model.

@@

 It is known that \acp{GNN} have a certain ability to generalise to new structures and scale with different nodes~\cite{DBLP:journals/aiopen/ZhouCHZYLWLS20,DBLP:conf/nips/KnyazevTA19}. 

@@

  Every $C$ steps, update the target network weights: $\theta^{-} \leftarrow \theta$
  ;
  }
  \caption{Deep Q-Network (DQN) with GNN and Graph Replay Buffer}
  \label{alg:dqn}
  \end{algorithm}

@@

This means that the average distance will tend towards the one expressed by the viewing range of each node:

@@

 The feature set used by the \ac{GNN} created for each node consists of the vector computed by the aggregated program and the value of the local sensor $\zeta$.

@@

\subsection{Baselines}
We compare our field-informed reinforcement learning approach against a baseline approaches where the \ac{DQN} utilises a \ac{MLP} as well as an approach only relying on \acp{GNN}, without additional field information. 
%
In all approaches, the underlying neural network (i.e., the \ac{MLP} and the \ac{GNN}) are trained %with the same parameters of the proposed approach 
with a single, stationary phenomenon.
%

>>>>

\and
%\linebreakand %\and
\IEEEauthorblockN{Mirko Viroli}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
Alma Mater Studiorum — Università di Bologna}\\
Cesena, Italy\\
mirko.viroli@unibo.it % 0000−0003−2702−5702
}
\and
\IEEEauthorblockN{Lukas Esterle}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
Aarhus University}\\
Aarhus, Denmark\\
lukas.esterle@ece.au.dk}
}
%
\maketitle              % typeset the header of the contribution
%
\input{scala.tex}
\lstset{language={scafi}}
%%% Comment command, to be removed before submission
\newcommand{\meta}[3]{\textcolor{#1}{\textbf{#2}: #3}}
\newcommand{\ga}[1]{\meta{red}{GA}{#1}}
\newcommand{\lukas}[1]{\meta{purple}{Lukas}{#1}}
\newcommand{\mv}[1]{\meta{green}{MV}{#1}}
%% ACRONYMS
\acrodef{DecPOMDP}[DecPOMDP]{decentralized partially observable Markov decision processes}
\acrodef{RL}[RL]{reinforcement learning}
\acrodef{MARL}[MARL]{multi-agent reinforcement learning}
\acrodef{MAARL}[ManyRL]{many-agent reinforcement learning}
\acrodef{MDP}[MDP]{Markov Decision Process}
\acrodef{MLP}[MLP]{multi-layer perceptron}
\acrodef{GNN}[GNN]{graph neural network}
\acrodef{CNN}[CNN]{convolutional neural network}
\acrodef{CTDE}[CTDE]{centralized training and decentralized execution}
\acrodef{DQN}[DQN]{Deep Q-Learning}
\acrodef{FIRL}[FIRL]{Field-Informed Reinforcement Learning}
%\ga{Page limit: 10 pages (include refs)}

@@

In this context, two main approaches have been proposed for creating distributed controllers out of macro-level task/goal descriptions: 
 \emph{manual design}, in which programmers build the controllers directly, and 
 \emph{automatic design}, which involves synthesizing programs using machine learning methods.
%
In this paper, we consider a new \emph{hybrid} approach called  \emph{Field-Informed reinforcement learning} (FIRL). We utilise manually designed \emph{computational fields} (globally distributed data structures) to manage the interaction between agents. Using \emph{Deep Q-learning} in combination with \emph{Graph Neural Networks} we enable the agents to learn the necessary local behaviour automatically to solve collective tasks. 
%
This allows us to create distributed controllers informed by a collective knowledge 
 that has been distilled during learning, but that uses only local information at runtime.
% 
We demonstrate the effectiveness of this new approach in simulated use cases 
 where tracking and covering tasks for swarm robotics are successfully solved. 
\end{abstract}

@@

%This leads to several questions 
%\begin{enumerate}
%	\item How to distribute sensors to ensure good attainment of information?
%	\item How to maintain knowledge about the phenomena and unlearn irrelevant information?
%	\item How to move sensors when the phenomena moves to ensure good coverage?
%\end{enumerate} \lukas{refine and rework these questions}
The coordination of a group of autonomous agents that can perceive and act in their environment is a fundamental problem in artificial intelligence. 
Such agents need to cooperate and communicate in order to achieve a common goal, 
 while dealing with the challenges of distributed and situated intelligence. 
 These challenges include the limited and local nature of the information available to each agent and the emergence of global behaviour from local interactions.

One domain where these challenges are particularly evident is Cyber-Physical Swarms (CPSs, or \emph{swarm-like systems}), 
 where a large number of agents interact with each other and their environment based on spatial proximity. 
%
Examples of CPSs include swarm robotics, smart cities~\cite{bajovic2021marvel}, sensor networks~\cite{pianini2022collective}, and social systems. 
 In these systems, agents need to adapt to dynamic and uncertain situations, while ensuring the achievement of global objectives that may not be directly observable or measurable by individual agents.

A key question in designing swarm-like systems is how to create distributed controllers for the agents that enable them to perform complex collective tasks. 
 Two main approaches have been proposed for this problem: manual design and automatic design. 
%
In manual design, programmers build the controllers directly, 
 using domain knowledge and programming languages or frameworks that support distributed computation and communication like macro-programming approaches~\cite{DBLP:journals/corr/abs-2201-03473}. 
%
In automatic design, machine learning methods like \ac{MARL} and evolutionary algorithms are used to synthesize programs or policies for the agents from the high-level task or goal descriptions.
%
Both approaches have advantages and disadvantages. 
 Manual design allows programmers to specify desired properties for the system declaratively. However, manual design can be tedious and error-prone.
%
Automatic design can overcome these limitations by learning from data and experience. 
 However, automatic design can also suffer from several challenges, such as learning the right representation for agent communication.
%

In this paper, we propose a novel hybrid approach that combines manual and automatic design of distributed controllers.
This follows a novel trend in which high-level declarative programming languages are mixed with machine learning techniques in order to synthesise robust collective controllers~\cite{DBLP:conf/acsos/Aguzzi21,DBLP:conf/icdcs/AguzziCV22,DBLP:conf/coordination/AguzziCV22}.
Specifically, we propose a solution called \ac{FIRL} that utilises aggregate computing~\cite{Beal2015Computer} together with a \ac{GNN}~\cite{Zhou2020AIOpen} in combination with a reinforcement learning approach, namely \ac{DQN}~\cite{mnih2015human}. 
% First, aggregate computing is utilised to collect information from neighbours and create a local knowledge base for each agent. Second, we utilise a \ac{GNN} within the \ac{DQN}. 
 Here the \ac{GNN} is trained on field-information from aggregate computing and provides so-called node embeddings for each agent, serving as input for the \ac{DQN}. 
 The \ac{DQN} provides then the appropriate actions for the agent to achieve their tasks.
%We further combine this with reinforcement learning techniques to respond to changing conditions in the environment. 
%
The main contribution of \ac{FIRL} is that it allows us to create distributed controllers that are informed by a collective knowledge that has been distilled during training, but that use only local information when deployed. This way, \ac{FIRL} can achieve a balance between manual design and automatic design, combining the benefits of both approaches while mitigating their drawbacks.
%% Perhaps to much detail 
\ac{FIRL} can also be seen as a way of bridging the gap between symbolic and sub-symbolic AI methods, by integrating declarative programming with deep learning.

We employ this approach in a swarm-like system setting where agents are tasked to cover phenomena detected in their environment. Over time, the agents have to converge over the each phenomenon in order to cover it appropriately as illustrated in \Cref{fig:simulations}.
\begin{figure}[t]
	\centering
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/start.png}
		\caption{Start}
		\label{fig:initial}
	\end{subfigure}
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/after.png}
		\caption{After 50 steps}
		\label{fig:after}
	\end{subfigure}
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/end.png}
		\caption{After 200 steps}
		\label{fig:end}
	\end{subfigure}
	\caption{Qualitative test results of the proposed approach. 
		The agents go towards the phenomenon and try to cover it without collapsing into a single point.
		the grey area represents the area covered by the agents (last figure on the right).}
	\label{fig:simulations}
\end{figure}

@@

\section{Background and Motivation}
\label{sec:background}
%In this section, we aim to provide a concise overview of the system we intend to address, 
% starting with a high-level system model. 
%%
%We then proceed to discuss relevant research areas that inform our approach. 
%%
%Finally, we present a formalization of the problem and a discussion of the motivations behind our proposed work.
%\ga{Page budget: 1/2 page \\}
%\ga{Plan: discussion about the problem of coordination in multi-agent systems and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{Swarm systems}
This article studies intelligent \emph{collective behaviours} within the context of \emph{large-scale} distributed systems.
 Specifically, we focus on \emph{Cyber-Physical Swarms} or \emph{swarm-like systems} consisting of \emph{devices} governed by autonomous software \emph{agents} and equipped with \emph{sensors} and \emph{actuator} able to interact with the real world.
 Each of these agents can interact with its direct neighbours, either based on physical (e.g. communication range) or logical distance (e.g. 1-hop neighbourhood).

%
Each agent
 executes a \emph{local} control loop. 
 In each iteration of this loop, the agent can access the information available within its own \emph{context}. 
 This context is comprised of information acquired by the agent directly through its sensors and the information received from its neighbours. 
%
At the end of each control loop iteration, 
 the agent can send messages to its neighbourhood. 
 This message may contain raw sensor data or aggregated information over time or from other neighbours.
% about its current evaluation and update its internal state.
%
%This general model can be adapted to several different scenarios, such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
Examples of such large-scale systems can be a network of robots (e.g., swarm robotics), camera and IoT networks, or a network of mobile phones.
%
Our goal is to find a \emph{homogeneously} distributed controller $\pi$ (i.e., the same controller for the whole system) that, 
 starting from \emph{only} local configurations, leads 
 the system to achieve a certain \emph{collective} requirements through \emph{cooperation}, 
 such as spatial area coverage, phenomena tracking, and robot aggregation~\cite{DBLP:journals/firai/SchranzUSE20}.
%
The homogeneity of the controller is a key requirement to ensure the \emph{scalability} of the approach, 
 as it allows us to avoid the need for a centralised controller that would be a bottleneck for the system,
 and it is the typical choice in swarm-like systems~\cite{brambilla2013swarm,yang2021many,pmlr-v80-yang18d,DBLP:conf/aaai/ZhengYCZZWY18}.
%
\subsection{Field-based Coordination}
%\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
The field-based coordination approaches utilise a concept of \emph{computational fields} 
 (or simply \emph{fields}), 
 which are \emph{distributed} data structures that associate locations with values and evolve over time. 
%
\emph{Aggregate computing}~\cite{Beal2015Computer}, a modern field-coordination approach, is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 
% 
This \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm can be realised to devise collective and self-organizing behaviour through a composition of functions operating on fields. 
%
The fields map a set of individual agents to computational values, 
 allowing them to associate what they sense, process, and actuate. 
%
Fields are computed locally but subject to a global viewpoint, 
 enabling emergent collective behaviour through the interplay of the \emph{system model} (i.e., structure and dynamics of the systems)
 and the \emph{programming model} (i.e., how collective behaviours are expressed) known as field calculus~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}.
%
The system model is structurally similar to \emph{swarm-like systems}, 
 as it is composed of a set of \emph{agents} 
 that interact with each other through a \emph{network} of \emph{neighbourhood relations}.
%
The dynamics of each agent follow a local control loop called \emph{round}
 that comprises the following steps:
\begin{enumerate}
  \item \emph{Sense}: the agent acquires information from the environment and collects the messages of its neighbourhood, building a \emph{context};
  \item \emph{Compute}: the agent executes a \emph{program} that computes an \emph{export}, that is a set of values to be sent to the neighbourhood;
  \item \emph{Act}: the agent sends the export to its neighbourhood and updates its internal state acting on the environment.
\end{enumerate}
The proactive and iterative execution of the round by each agent leads to the emergence of collective behaviour specified by the program.

@@

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source agent and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. 
%
In ScaFi~\cite{casadei2022scafi} %(a scala implementation of aggregate computing)
 it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the field expansion. 
This pattern is called \emph{gradient-cast} (or \emph{G}) and is a fundamental building block for the coordination of swarm-like systems.
It is expressed as: 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
%\ga{more description here}
where \texttt{source} is a Boolean field indicating the presence of a source agent, 
 \texttt{value} is the value to be cast, and 
 \texttt{acc} is the accumulation function that is applied to the value during the cast.

%\lukas{Lukas to review this section again.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $f_v$.
For the sake of simplicity, 
 we thereafter describe $G_f$ as a graph that contains the feature set $f_v$ for each node $v \in V$.
 Note that, when we refer to $G_f$ and $G_o$ we are referring to the same graph $G$ but with different node features.
 Also, to access the feature set $f_v$ of a node $v \in V$ we use the notation $f_v$ or $G_f[v]$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $f_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. 
 The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
%
In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $\mathcal{N}_G(v)$,
 and then combining it with the node's current embedding $h_v$.
%
The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. 
 With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}_G(v)\right\}\right)   
\end{equation}

@@

	\centering
  \includegraphics[width=.98\linewidth]{imgs/architecture.pdf}

@@

  \caption{High-level description of \ac{FIRL} approach. 

@@

  Using this feature, an aggregate program computes spatio-temporal information that enhances the feature set of each agent, depicted as colours in the middle graph. 
  Finally, utilizing the GNN, actions are computed for each agent in the system to be performed against the environment, enabling advancement in the simulations according to swarMDP rules.

@@

 representing the participating agents and their neighbourhood relations at time $t$, as input. 

@@

The policy $\pi_{gnn}$ is then evaluated for each agent using $f_v$ as input, 

@@

The individual agents only combine information from their local neighbourhood.

@@

In fact, the program $\Gamma$ is proactively executed at every agent, 

@@

otherwise, they could not have a local interpretation for each agent, according to our system model.

@@

 It is known that \acp{GNN} have a certain ability to generalise to new structures and scale with different agents~\cite{DBLP:journals/aiopen/ZhouCHZYLWLS20,DBLP:conf/nips/KnyazevTA19}. 

@@

  Every $C$ steps, update the target network weights: $\theta^{-} \leftarrow \theta$
  ;
  }
  \caption{Deep Q-Network (DQN) with GNN and Graph Replay Buffer executed by each agent}
  \label{alg:dqn}
  \end{algorithm}

@@

This means that the average distance will tend towards the one expressed by the viewing range of each agent:

@@

 The feature set used by the \ac{GNN} created for each agent consists of the vector computed by the aggregated program and the value of the local sensor $\zeta$.

@@

\subsection{Baselines}
We compare our \ac{FIRL} approach against a baseline approaches where the \ac{DQN} utilises a \ac{MLP} as well as an approach only relying on \acp{GNN}, without additional field information. 
%
In all approaches, the underlying neural network (i.e., the \ac{MLP} and the \ac{GNN}) are trained %with the same parameters of the proposed approach 
with a single, stationary phenomenon.
%

The \ac{MLP} uses the same feature set as the \ac{GNN} but applies it in the \ac{DQN} but 
%in the proposed approach, but differently, it uses a standard DQN approach 
without leveraging the graph structure. Moreover, we increase the batch size to 512 and the replay buffer to 10000 since we record 25 agents' experiences for each step instead of one graph experience.
%\lukas{to make sure: in all three cases we use DQN but with (i) MLP, (ii) standard GNN, (iii) our approach - a field-informed GNN, right?}
%
%The \ac{GNN} instead of using the field information, use directly the position of nodes and the local sensor value as input features.
The \ac{GNN} alone, without using the field information, apply the position of agents and the local sensor value directly as input features within the \ac{DQN}.
%
These baselines are used to verify the effectiveness of the components used in the field-informed \ac{RL}. 
%
Indeed the \ac{MLP} baseline is used to verify the effectiveness of the \ac{GNN} in the proposed approach, while the \ac{GNN} baseline is used to verify the effectiveness of the field information in the proposed approach.
\subsection{Metrics}
We evaluate the performance of the different approaches by measuring the coverage of the phenomenon over time.
%
The coverage is defined as the percentage of the phenomenon covered by the agents.
%
Specifically, we can measure the coverage as the intersection over the union of the phenomenon and the agents' view range.
%
Formally, first, we define the overall coverage for a certain time step as:
\begin{equation*}
\Omega = \bigcup_{v \in V} \omega_v
\quad 
C = \frac{|\Omega \cap \mathcal{P}|}{|\mathcal{P}|}
\end{equation*}
where $\mathcal{P}$ is the area of the phenomenon and $\Omega$ is the area covered by the agents.
%This is computed for each step of the simulation. 
In the training phases, 
 we measure the average coverage in each episode, and the total reward obtained by the agents at each episode of the simulation. 
%
We also measure the number of agents that are within the phenomenon at each step of the simulation.
This will be a measure of how well the agents are tracking the phenomenon.
\ga{todo improve the description of the coverage}
